The global implications of taxing land in agricultural production: Land use, agricultural production, and consumer well-being

### 2024-03-14 paper.Rmd
Update paper.Rmd

---
title: "Avoiding global deforestation by taxing land in agricultural production: The implications for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected global growth in population and incomes is expected to create pressure to convert forestland into farmland, with an estimated 144.2 million hectares of forestland estimated to be converted to agricultural land by 2033. At the same time, increasingly negative climate impacts are expected to generate further pressure to enhance the terrestrial carbon sink, but taking such actions would likely result in less agricultural output and higher food prices. Using the GTAP and GTM models, we model how different tax rates on the value of output from land used in agricultural production would impact deforestation. We find that, with a tax rate of zero percent on agricultural production, 144.2 million hectares (3.6%) of global forests would be converted to agricultural lands by 2033. With a 70 percent tax, however, the net deforestation rate would be brought to zero while the well-being of global consumers would be reduced by $92.1 billion in 2033, which translates to a global average annual cost of $10.6 per person."
---

@@

Our second scenario examines the impact of a tax of roughly 70 percent on the use of agricultural land. Solving the two models together shows that the economic impact would not be limited to the cost of the tax, it would also impact food production and its prices. For example, most regions of the world would see significant declines in the amount of agricultural land in 2033 relative to the baseline scenario where land conversion was not taxed. For the United States, there would be 2.1 percent less agricultural land (Table 3). Larger decreases would happen in other major agricultural production regions such as China (-2.5 percent), South Asia (-7.9 percent), Brazil (-8.6 percent), and Russia (-9.3 percent). This would be because the global net deforestation rate would be brought to zero and 144.2 million hectares that would otherwise have been converted to agricultural land by 2033 would be preserved as forestland.

These reductions in the amount of agricultural land would then impact the quantity of agricultural goods that could be produced. For example, the 15.0 percent decrease in agricultural land in East Asia and the 9.0 percent decrease in agricultural land in Southeast Asia would result in reductions in agricultural production of 5.0 percent and 1.7 percent, respectively. Surprisingly, Brazil and the United States, despite their decreases in available agricultural land compared to the baseline 2033 scenario, are estimated to see increased agricultural output of 0.6 percent and 1.0 percent, respectively. This is the result of the sharp output declines in other regions prompting increased intensity of production in the United States and Brazil. When we examine how the policy impacts exports of agricultural goods, we find that exports from Japan and the United States grow the most, increasing 8.8 percent and 3.5 percent, respectively (Table 4).

@@

Overall, the 70 percent tax, which results in zero net deforestation globally, is liekly to have a negative economic impact on all regions except Japan and Canada, which see an increase in consumer well-being of
$977.1 million and $166.9 million, respectively (Table 5). Consumer well-being is a measurement of equivalent variation and here measures the benefit/harm consumers experience in terms of income as a result of the policy changes. The United States, along with 10 other regions, are estimated to see well-being drop by more than $1 billion. The most negatively impacted regions would be South Asia (-$28.0 billion), China (-$14.5 billion), and Southeast Asia (-$13.6 billion). In total, this tax-driven policy is estimated to decrease global consumer well-being in 2033 by $92.1 billion relative to the scenario where no tax on agricultural land was in place.

@@

This analysis shows that, at an estimated global tax rate of 70.3 percent of the value of output from land used in agricultural production, global net deforestation would drop to roughly zero by 2033, preserving 144.2 million hectares of forestland that otherwise would have been converted to agricultural land. The economic price of this would be transferred to consumers through lower agricultural output and higher market prices. Overall, this land tax would reduce consumer well-being by $92.1 billion in 2033, which equates to $10.60 per person annually. Governments might be wary of farmer discontent were they to take such fiscal actions and thus might be reluctant to impose such taxes on them. Thus, if policymakers wish to see deforestation stopped, they could instead use import tariffs on products produced on land converted from forestland for agricultural use. Those scenarios, however, will be covered in a subsequent study.

@@

According to CEPII’s projections, the growth of the global economy in the next ten years is expected to result in tremendous increases in global and regional GDP accompanied by a moderate growth in population. Another set of projections by GTM suggests that this growth would also mean significant reductions in the world’s forests because they would be converted to agricultural land, with the model predicting a loss of over 144 million hectares of forest by 2033.

By combining two well-known global models—the GTAP model that describes the world markets and trade, and the GTM model that describes the impact of economic and population growth along with land rents on the amounts of forested areas—we were able to estimate the size of the tax on agricultural land needed to

prevent deforestation in the coming ten years through reducing the returns to land when used in agriculture. We find that even though the tax is substantial at about 70 percent, it results in only very small reductions in the projected growth in global GDP and welfare. We also observe that the reduction in available agricultural land relative to the baseline growth results in reduced agricultural output and higher prices, which may require additional policy action to avoid negative impacts in terms of food security.

Finally, we note that a uniform global tax reducing the returns to land to the point of preventing global deforestation would produce highly differentiated impacts around the world’s regions. Some regions, such as China, Russia, Canada, and the United States would increase their forested areas as a result of the tax, while the regions with the greatest deforestation pressures, such as Sub-Saharan Africa, Brazil, and Southeast Asia would only reduce the rate of their deforestation, for a zero change in the global forest area. Therefore, future studies will need to consider how best to target policies to counteract deforestation in zones facing the greatest land use conversion pressures.

>>>>

---
title: "The global implications of taxing land in agricultural production: Land use, agricultural production, and consumer well-being"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected global growth in population and incomes is expected to create pressure to convert forestland into farmland, with an estimated 144.2 million hectares of forestland estimated to be converted to agricultural land by 2033. At the same time, increasingly negative climate impacts are expected to generate further pressure to enhance the terrestrial carbon sink, but taking such actions would likely result in less agricultural output and higher food prices. Using the GTAP and GTM models, we model how different tax rates on the value of output from land used in agricultural production would impact deforestation. We find that, with a tax rate of zero percent on agricultural production, 144.2 million hectares (3.6%) of global forests would be converted to agricultural lands by 2033. With a 70 percent tax, however, the net deforestation rate would be brought to zero while the well-being of global consumers would be reduced by $92.1 billion in 2033, which translates to a global average annual cost of $10.6 per person."
---

@@

Our second scenario examines the impact of a tax of roughly 70 percent on the use of agricultural land. Solving the two models together shows that the economic impact would not be limited to the cost of the tax, it would also impact food production and its prices. For example, most regions of the world would see significant declines in the amount of agricultural land in 2033 relative to the baseline scenario where land conversion was not taxed. For the United States, there would be 2.1 percent less agricultural land (Table 3). Larger decreases would happen in other major agricultural production regions such as China (-2.5 percent), South Asia (-7.9 percent), Brazil (-8.6 percent), and Russia (-9.3 percent). This would be because the global net deforestation rate would be brought to roughly zero, meaning that the 144.2 million hectares that would have been converted to agricultural land by 2033 in the zero tax scenario would be preserved as forestland.

This would have impacts on the quantity of agricultural goods that could be produced. For example, the 15.0 percent decrease in agricultural land in East Asia and the 9.0 percent decrease in agricultural land in Southeast Asia would result in reductions in agricultural production of 5.0 percent and 1.7 percent, respectively. Surprisingly, Brazil and the United States, despite their decreases in available agricultural land compared to the baseline 2033 scenario, are estimated to see increased agricultural output of 0.6 percent and 1.0 percent, respectively. This is the result of the sharp output declines in other regions prompting increased intensity of production in the United States and Brazil. When we examine how the policy impacts exports of agricultural goods, we find that exports from Japan and the United States grow the most, increasing 8.8 percent and 3.5 percent, respectively (Table 4).

@@

Overall, the 70 percent tax, which results in zero net deforestation globally, is likely to have a negative economic impact on all regions except Japan and Canada, which see an increase in consumer well-being of
$977.1 million and $166.9 million, respectively (Table 5). Consumer well-being is a measurement of equivalent variation and here measures the benefit/harm consumers experience in terms of income as a result of the policy changes. The United States, along with 10 other regions, are estimated to see well-being drop by more than $1 billion. The most negatively impacted regions would be South Asia (-$28.0 billion), China (-$14.5 billion), and Southeast Asia (-$13.6 billion). In total, this tax-driven policy is estimated to decrease global consumer well-being in 2033 by $92.1 billion relative to the scenario where no tax on agricultural land was in place.

@@

This analysis shows that tax rates on agricultural land can have a significant impact on land use. At a global average tax rate of zero percent of the value of land used in agricultural production, an estimated 144.2 million hectares of forestland would be converted to agricultural land by 2033. With an estimated global tax rate of 70.3 percent, global net deforestation would drop to roughly zero by 2033, preserving 144.2 million hectares of forestland. The economic price of this would likely be transferred to consumers through lower agricultural output and higher market prices. Overall, this land tax is estimated to reduce consumer well-being by $92.1 billion in 2033, which equates to $10.60 per person annually.

@@

According to CEPII’s projections, the growth of the global economy in the next ten years is expected to result in tremendous increases in global and regional GDP accompanied by a moderate growth in population. Another set of projections by GTM suggests that this growth would also mean significant reductions in the world’s forests because they would be converted to agricultural land, with the model predicting a loss of over 144 million hectares of forest by 2033. By combining two well-known global models—the GTAP model that describes the world markets and trade, and the GTM model that describes the impact of economic and population growth along with land rents on the amounts of forested areas—we were able to estimate the impact of a tax on agricultural land on land use change. We find that a tax of about 70 percent would likely prevent net global deforestation in the coming ten years through reducing the returns to land when used in agriculture. We also observe that the reduction in available agricultural land relative to the baseline growth would likely result in reduced agricultural output and higher prices, which might have food security implications. Finally, we note that such a uniform global tax that would reduce the returns to land to the point of preventing global deforestation appears to produce highly differentiated impacts around the world’s regions. Some regions, such as China, Russia, Canada, and the United States are estimated to see their forested areas increase as a result of the tax, while the regions with the greatest deforestation pressures, such as Sub-Saharan Africa, Brazil, and Southeast Asia would only reduce the rate of their deforestation, for a zero change in the global forest area.
